---
title: '#HEALTH INSURANCE FRAUD DETECTION'
author: "Mohammed Ba-Aoum"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
```


```{r}
# Load the dataset
file_path <- "Health_Insurance_Fraud.xlsx"
data <- read_excel(file_path, sheet = "Fraud_Detection_decsion tree")

```



```{r}
# Explore the data
# Clean column names (e.g., replace invalid ones)
names(data) <- make.names(names(data))
str(data)
summary(data)
```


```{r}
# Convert relevant columns to factors
factor_vars <- c("fraud_reported", "police_report_available", 
                 "policy_state", "auto_make", "auto_model", "policy_csl")
data[factor_vars] <- lapply(data[factor_vars], as.factor)

```


```{r}
# Replace "?" with NA only in character columns
char_cols <- sapply(data, is.character)
data[char_cols] <- lapply(data[char_cols], function(x) ifelse(x == "?", NA, x))
data <- na.omit(data)  # Drop rows with missing values

```


```{r}
# Drop unnecessary columns

data_clean <- data %>%
  select(-c(policy_number, policy_bind_date, insured_zip, incident_location))
```


# EDA Visualizations 
```{r}
# 1. Fraud report distribution
ggplot(data_clean, aes(x = fraud_reported, fill = fraud_reported)) +
  geom_bar() +
  labs(title = "Fraud Reported Distribution", x = "Fraud Reported", y = "Count")
```
#Insight:
This plot shows that a majority of the insurance claims in the dataset are non-fraudulent (N), but a significant proportion (around 25%) are reported as fraudulent (Y). This class imbalance needs to be kept in mind when modeling, as it may impact model sensitivity and specificity.

```{r}
# 2. Incident severity by fraud status
ggplot(data_clean, aes(x = incident_severity, fill = fraud_reported)) +
  geom_bar(position = "dodge") +
  labs(title = "Incident Severity vs Fraud Reported", x = "Incident Severity")


```
# Insight:
Claims involving "Major Damage" are more frequently reported as fraudulent compared to those with "Minor" or "Trivial" damage. This suggests that claim severity is a strong indicator of potential fraud, likely because larger claims present a higher opportunity for abuse.

```{r}
# 3. Boxplot of total claim amount by fraud
ggplot(data_clean, aes(x = fraud_reported, y = total_claim_amount, fill = fraud_reported)) +
  geom_boxplot() +
  labs(title = "Total Claim Amount by Fraud Status", y = "Total Claim Amount")


```
# Insight:
The boxplot reveals that fraudulent claims tend to have a wider range and higher median total claim amounts. This supports the intuition that fraudsters often inflate claim values to maximize gain, making this feature particularly valuable for prediction.

```{r}
# Split data into training and testing sets
set.seed(123)
trainIndex <- createDataPartition(data_clean$fraud_reported, p = 0.7, list = FALSE)
train <- data_clean[trainIndex, ]
test <- data_clean[-trainIndex, ]


```

```{r}

# Build a Decision Tree model
tree_model <- rpart(fraud_reported ~ ., data = train, method = "class")
rpart.plot(tree_model)

```
```{r}


# Predict on test data
predictions <- predict(tree_model, test, type = "class")

```


```{r}
# Evaluate model performance
confusionMatrix(predictions, test$fraud_reported)


```


```{r}
# Random Forst

rf_model <- randomForest(fraud_reported ~ ., data = train, ntree = 100, importance = TRUE)
rf_pred <- predict(rf_model, test)
confusionMatrix(rf_pred, test$fraud_reported)
```
```{r}
#Feature Importance
# Extract variable importance
importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)
importance_df %>%
  arrange(desc(MeanDecreaseGini)) %>%
  top_n(20, MeanDecreaseGini) %>%
  ggplot(aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 20 Important Features (Random Forest)", 
       x = "Variable", y = "Mean Decrease in Gini") +
  theme_minimal()
```
# Insight:
The random forest model identifies features like incident_severity, auto_model, auto_make, insured_hobbies, and property_claim as some of the most important for detecting fraud. These variables offer valuable behavioral and contextual cues, showing that not only financial metrics but also customer behavior patterns contribute to effective fraud detection.